---
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Analysis on NBA players statistical performance
#### Dominic DiRe, Blake Lewis, Will Stone

## Introduction

This project explores NBA player and team statistics pulled from the NBA API across 5 seasons (2020-21 season through 2024-25 season). Basketball is a sport with a great tie to statistics, and modern data collection allows us to track how individual players and the league as a whole evolve over time, although it's just 5 seasons. Analysis on this topic can reveal trends in player development, positional changes in the modern NBA, and how elite players separate themselves from the rest of the league. The goal is to visualize league-wide stat distributions and track individual player development over time, with a focus on Luka Doncic, the future face of the NBA. Who we believe might be able to be a good basis to separate the stars from the future hall of famers.

In pursuit of the goal stated above, we will explore the following questions:

1. What does the distribution of key stats (points, rebounds, assists) look like across the league, and how have those distributions shifted across seasons?

2. Which players are the most efficient scorers? How does scoring volume relate to shooting efficiency?

3. Who are the top 10 scorers, rebounders, and assisters each season, and how much turnover is there in those lists year over year?

4. How has the league's reliance on the three-point shot changed across the 5 seasons?

5. Is there a relationship between minutes played and points per game?

6. Which players have the best plus/minus relative to their scoring load — who is impacting winning beyond just putting up points?

7. Which players improved or regressed the most in scoring?

8. How has Luka Doncic developed from season to season across points, assists, rebounds, and shooting percentages?

9. How do Luka's per-game stats compare to league averages each season? How about compared to the MVP's of those seasons?

These are the main questions we are looking to answer through the completion of this project. With the findings we will be able to draw meaningful conclusions on player development and league-wide trends in the NBA.

## Data

### Structure

Data was pulled from the NBA Stats API (`stats.nba.com`) using the `nba_api` Python package. The python file was basic but too long and embarrassing to add, if divine intervention strikes we can add it. Player game logs were collected for every regular season game across 5 seasons and aggregated into per-game averages. This produced one CSV per season for league-wide stats.Also there is another separate combined CSV for Luka Doncic spanning all 5 seasons.

**League CSVs** (`league_stats_YYYY-YY.csv`) — one row per player, containing per-game averages for all NBA players in that regular season.

**Luka CSV** (`luka_career_stats.csv`) — one row per season, pulled from the PlayerCareerStats endpoint.

**Columns included:** `PLAYER_ID`, `PLAYER_NAME`, `TEAM_ABBREVIATION`, `GP`, `MIN`, `PTS`, `REB`, `AST`, `STL`, `BLK`, `FG_PCT`, `FG3_PCT`, `FT_PCT`, `TOV`, `PLUS_MINUS`

### Loading

```{r, warning=FALSE, message=FALSE}
library(tidyverse)

league_2021 <- read_csv("league_stats_2020-21.csv")
league_2022 <- read_csv("league_stats_2021-22.csv")
league_2023 <- read_csv("league_stats_2022-23.csv")
league_2024 <- read_csv("league_stats_2023-24.csv")
league_2025 <- read_csv("league_stats_2024-25.csv")

luka <- read_csv("luka_career_stats.csv")
```

### Merging

Since each league CSV covers a single season but contains no season identifier, so we added a `SEASON` column to each dataframe before combining them. Without the column there wouldn't be a way to tell where data came from

```{r}
league_2021$SEASON <- "2020-21"
league_2022$SEASON <- "2021-22"
league_2023$SEASON <- "2022-23"
league_2024$SEASON <- "2023-24"
league_2025$SEASON <- "2024-25"
```

With the season identifier in place, we combined all 5 dataframes into a single `df1` (later is `master`) dataframe using `bind_rows`. `left_join` merges datasets horizontally but for this dataset that doesn't make as much sense. So we used `bind_rows`, which stacks datasets vertically. This is appropriate here because all 5 league CSVs share identical columns — they are the same data structure repeated across different seasons. Which makes this process quite easy.
```{r}
df1 <- bind_rows(league_2021, league_2022, league_2023, league_2024, league_2025)
```

The raw data includes all players who appeared in at least one regular season game, including players who played only a handful of minutes across a few games. We filtered to players with at least 20 games played to ensure per-game averages are based on a meaningful sample size. Also if we didn't add this we had a player who scored 30 points in a single game appearance would show a 30.0 PPG average and distort league-wide distributions. Also running this line puts the number of observations from 3261 to 2316, so the change it quite noticeable.
```{r}
df1 <- df1 %>% filter(GP >= 20)
```

To ease running the code, we saved the merged and slightly cleaned dataset to "master.csv" so you only have to run this line. I know we aren't at all going for efficiency this project but I still feel like it's nice to have.
```{r, warning=FALSE, message=FALSE}
write.csv(df1, "master.csv", row.names = FALSE)
```


### Variables

* PLAYER_ID: Unique numeric identifier for each player assigned by the NBA.
* PLAYER_NAME: First Last style full name of player.
* TEAM_ABBREVIATION: The 3-letter abbreviation of the team the player played for (e.g. DAL, LAL).
* SEASON: The NBA season the stats were recorded in, in YYYY-YY format (e.g. 2023-24).
* GP: Games Played, the number of regular season games not including any play-ins or playoffs.
* MIN: Minutes Per Game, the average minutes for that player per game.
* PTS: Points Per Game, the average points for that player per game.
* REB: Rebounds Per Game, the average number of total rebounds (offensive and defensive grouped) per game.
* AST: Assists Per Game, the average assists for that player per game.
* STL: Steals Per Game, the average steals for that player per game.
* BLK: Blocks Per Game, the average blocks for that player per game.
* FG_PCT: Field Goal Percentage, the share of all field goal attempts that were made by player, expressed as a decimal between 0 and 1.
* FG3_PCT: Three-Point Percentage, the share of three-point attempts that were made by player, expressed as a decimal between 0 and 1.
* FT_PCT: Free Throw Percentage, the share of free throw attempts that were made by player, expressed as a decimal between 0 and 1.
* TOV: Turnovers Per Game, the average turnovers for that player per game.
* PLUS_MINUS: The average point differential for the player's team while they were on the court.

### Cleaning

First, read the master.
```{r, warning=FALSE,message=FALSE}
library(readr)
library(tidyverse)
master <- read_csv('master.csv')
```

Problem with the dataset as is, is that the traded players or players who get picked up by multiple team in a season, the player has 2 team rows. This caused a problem a bit later using `pivot_wider`, where it didn't know what to pick essentially. Now this function is to sort out players with multiple team rows in the same season.

```{r}
master <- master %>%
  group_by(PLAYER_ID, PLAYER_NAME, SEASON) %>%
  summarise(
    GP = sum(GP),
    MIN = mean(MIN),
    PTS = mean(PTS),
    REB = mean(REB),
    AST = mean(AST),
    STL = mean(STL),
    BLK = mean(BLK),
    FG_PCT = mean(FG_PCT, na.rm = TRUE),
    FG3_PCT = mean(FG3_PCT, na.rm = TRUE),
    FT_PCT = mean(FT_PCT, na.rm = TRUE),
    TOV = mean(TOV),
    PLUS_MINUS = mean(PLUS_MINUS),
    .groups = "drop"
  )
```


The `master` dataframe is in "long format", where each player has one row per season. This structure is wpuld be good for league-wide distributions and filtering by season. However, for comparing the same player's stats side by side across multiple seasons, we also opted for creating a "wide format" dataframe using `pivot_wider`. In `master_wide`, each player occupies a single row and each stat gets its own column per season, such as `PTS_2020-21` and `PTS_2021-22`. Players who did not appear in a given season will have `NA` for that season's columns, newer rookies and retirees.

```{r, warning=FALSE, message=FALSE}
master_wide <- master %>%
  select(PLAYER_ID, PLAYER_NAME, SEASON, PTS, REB, AST, STL, BLK,
         FG_PCT, FG3_PCT, FT_PCT, TOV, PLUS_MINUS) %>%
  pivot_wider(
    names_from = SEASON,
    values_from = c(PTS, REB, AST, STL, BLK, FG_PCT, FG3_PCT, FT_PCT, TOV, PLUS_MINUS),
    names_glue = "{.value}_{SEASON}"
  )
```

From `master_wide` we got a `PTS_growth` column by subtracting a player's 2020-21 scoring average from their 2024-25 average, this is work in progress so far but the idea is there. This gives us a simple stat to identify which players improved the most in scoring over the full 5 year span. Players with `NA` in either season, meaning they did not meet the games played threshold in one of those years, will have `NA` for `PTS_growth` as well.

```{r, warning=FALSE, message=FALSE}
master_wide <- master_wide %>%
  mutate(PTS_growth = `PTS_2024-25` - `PTS_2020-21`)
```

## Analysis

### What does the distribution of key stats look like across the league?

To kick off the exploration and analysis, the scoring distribution is probably what people would first think of showing. The scoring distribution of the most recent season will set a good sort-of baseline for the shape of the league in scoring. The distribution is right-skewed. Meaning the majority of players are role players and not stars, not surprising but notable. Then you can see the high volume scorers on the right

```{r}
master %>%
  filter(SEASON == "2024-25") %>%
  ggplot(aes(x = PTS)) +
  geom_histogram(binwidth = 1, fill = "blue", color = "white") +
  labs(title = "Distribution of points per game (ppg) — 2024-25",
       x = "Points per game (ppg)", y = "Number of players")
```

Now that a baseline is set for ppg, we are going to look at the 3 key stats over all 5 seasons. The 3 stats being points, assists, and rebounds. We use `facet_wrap` to compare the shapes of the points, rebounds, and assists distributions side by side. So you can see how production of the key stats is changed between the seasons.

Originally we tried to use a histogram, but wow was that a little hard to read. It looked nice and we wanted to keep consistency from the histogram from above, but we decided a boxplot is going to look much prettier. And I think we were right, any sane data science or stats major should be foaming at the mouth looking at these.

```{r}
master %>%
  select(PLAYER_NAME, SEASON, PTS, REB, AST) %>%
  pivot_longer(cols = c(PTS, REB, AST), names_to = "stat", values_to = "value") %>% # tech
  ggplot(aes(x = SEASON, y = value, fill = SEASON)) +
  geom_boxplot() +
  facet_wrap(~ stat, scales = "free_y") + # TECH, also so each avg diff y
  labs(title = "Distribution of points, assists, rebounds by season",
       x = "Season", y = "Per game averages", fill = "Season") +
  theme(axis.text.x = element_blank()) # was overcrowding, would be fine if no facet_wrap
```

### Which players are the most efficient scorers?

```{r}
# placeholder
```

### Who are the top 10 scorers, rebounders, and assisters each season?

```{r}
# placeholder
```

### How has the league's reliance on the three-point shot changed?

```{r}
# placeholder
```

### Is there a relationship between minutes played and points per game?

It's pretty natural to think that if a player gets more minutes, they will get more opportunity to score. We wanted to look if players had more minutes, they would generally score more than people who didn't, and take notes of outliers. We made a scatterplot that looks at all the seasons in our dataset with a trend line to see if minutes per game can actually predict points scored in a game. Players who score higher than 25 points per game will be labeled differently, marked as red dots. We originally labeled names but it just looks too cluttered, and if we forced them not to overlap a top of names went unseen.

```{r}
master %>%
  mutate(elite = PTS > 25) %>%
  ggplot(aes(x = MIN, y = PTS, color = elite)) +
  geom_point(alpha = 0.3) +
  scale_color_manual(values = c("FALSE" = "blue", "TRUE" = "red"),
                     labels = c("FALSE" = "Under 25 PPG", "TRUE" = "25+ PPG")) +
  geom_smooth(method = "lm", color = "black") + # for straight line
  labs(title = "Minutes per game vs points per game for all 2020-2025 seasons",
       x = "Minutes per game", y = "Points per game")

```
Another cool thing to look at is the specifics of the correlation of the two variables minutes and points. r = .87 which is a pretty strong correlation, as well as the p value being extremely small. So there is overwheling evidence that the minutes and points variables are correlated.
```{r}
cor_test <- cor.test(master$MIN, master$PTS)
cor_test
```

We can extend this further by looking at whether heavier minutes also relate to shooting efficiency. Trying to find if fatigue is really a problem or not. The scatterplot plots minutes per game against field goal percentage. Players with high minutes and high FG% are the most durable and efficient players in the league — a rare combination, and don't worry the Don is there somewhere.

```{r}
master %>%
  filter(FG_PCT > 0.1) %>%
  mutate(highlight = case_when(
    MIN > 30 & FG_PCT > 0.52 ~ "High Min, High FG%",
    MIN > 30 & FG_PCT < 0.42 ~ "High Min, Low FG%",
    TRUE ~ "Other"
  )) %>%
  ggplot(aes(x = MIN, y = FG_PCT, color = highlight)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  scale_color_manual(values = c(
    "Other" = "gray",
    "High Min, High FG%" = "purple",
    "High Min, Low FG%" = "green"
  )) +
  labs(title = "Minutes per Game vs FG%",
       x = "Minutes per game", y = "FG%", color = "")
```

```{r}
cor_test2 <- cor.test(master$MIN, master$FG_PCT)
cor_test2
```

### Which players have the best plus/minus relative to their scoring load?

Plus/minus is a player's way of impacting the game without their stats, so esentially how many points the team scores and lets up when they are on the floor. It's a great indicator to tell the story of players who don't score a ton, but are amazing on defense or playmaking, contributing greatly to their teams winning.The scatterplot below shows point per game against plus/minus for 2024-25. Horizontal base line at 0 to split the good and bad plus/minus. The upper left of the graph are players with less scoring, but good plus minus. And the top right is the true outliers of basketball. So this is essentially capturing players with a sort of hidden value.

```{r}
# install.packages("ggrepel")
library(ggrepel) # found easier package to label, might need to install to run

master %>%
  filter(SEASON == "2024-25") %>%
  mutate(highlight = case_when(
    PLUS_MINUS > 5 & PTS < 12 ~ "High PM, low scoring",
    PLUS_MINUS > 8 & PTS > 20 ~ "High PM, high scoring",
    TRUE ~ "Non-outliers" # default
  )) %>%
  ggplot(aes(x = PTS, y = PLUS_MINUS, color = highlight)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  geom_text_repel(
    data = . %>% filter(highlight != "Non-outliers"), # otherwise everyone highlighted
    aes(label = PLAYER_NAME),
    size = 2.5, show.legend = FALSE
  ) +
  scale_color_manual(values = c(
    "Non-outliers" = "violet",
    "High PM, low scoring" = "red",
    "High PM, high scoring" = "black"
  )) +
  labs(title = "Points per game vs Plus/Minus — 2024-25 season",
       x = "Points per game", y = "Plus/Minus")
```
To complement the single season view, this bar chart below shows the top 10 individual plus/minus performances across all 5 seasons with an asterisk. Each bar is colored by season, highlighting peak impact contributers. Jokic actually had 3 seasons that could make it on this graph, (22-23, 23-24, 24-25), but for sake of top 10 players we edited the conditions for making it using `with_ties`.

```{r}
master %>%
  group_by(PLAYER_NAME) %>%
  slice_max(PLUS_MINUS, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  slice_max(PLUS_MINUS, n = 10, with_ties = FALSE) %>%
  ggplot(aes(x = reorder(PLAYER_NAME, PLUS_MINUS), y = PLUS_MINUS, fill = SEASON)) +
  geom_col() +
  coord_flip() + # looks so much cooler like this
  labs(title = "Top 10 single season Plus/Minus performances",
       subtitle = "Across all 5 seasons",
       x = "", y = "Plus/Minus", fill = "Season")
```

### Which players improved or regressed the most in scoring?

Using the `master_wide` and the `PTS_growth` made weeks ago, we can identify the players that made the biggest jumps, and biggest drops in scoring from 2020 to 2025.
This only includes players that were in the 2020-21 seasons and the 2024-25 season. So it shows players that didn't just grow from 1 pt scored to 8 from 2023 season to the 2024 season and highlight them as the next coming of Jordan. But this also means that recent rookies and players that retired are excluded.

```{r}
master_wide %>%
  filter(!is.na(PTS_growth)) %>%
  slice_max(PTS_growth, n = 10) %>%
  ggplot(aes(x = reorder(PLAYER_NAME, PTS_growth), y = PTS_growth)) +
  geom_col(fill = "skyblue") +
  coord_flip() +
  labs(title = "Top 10 most improved scorers",
       subtitle = "2020-21 to 2024-25",
       x = "", y = "Points per game growth")
```

```{r}
master_wide %>%
  filter(!is.na(PTS_growth)) %>%
  slice_min(PTS_growth, n = 10) %>%
  ggplot(aes(x = reorder(PLAYER_NAME, -PTS_growth), y = PTS_growth)) +
  geom_col(fill = "tomato") +
  coord_flip() +
  labs(title = "Top 10 most regressed scorers",
       subtitle = "2020-21 to 2024-25",
       x = "", y = "Points per game growth")
```

The scatter below plots every player's 2020-21 scoring average against their 2024-25 average. Players above the diagonal line improved, players below it regressed. This tells a much better story than just the top and bottom 10, showing the full picture of how scoring output shifted across the league over 5 seasons and pointing out the outliers.

```{r}
master_wide %>%
  filter(!is.na(PTS_growth)) %>%
  ggplot(aes(x = `PTS_2020-21`, y = `PTS_2024-25`)) +
  geom_point(alpha = 0.4, color = "blue") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "black") + # sloped line
  geom_text(
    data = . %>% filter(PTS_growth > 10 | PTS_growth < -8),
    aes(label = PLAYER_NAME),
    size = 2.5, vjust = -0.5, check_overlap = TRUE
  ) +
  labs(title = "Scoring Average: 2020-21 vs 2024-25",
       subtitle = "Players above the line improved, below regressed",
       x = "Points per game 2020-21", y = "Points per game 2024-25")
```

### How has Luka Doncic developed season to season?

```{r}
# placeholder
```

### How do Luka's stats compare to league averages each season? How about compared to the MVP's of those seasons?

```{r}
# placeholder
```

## Conclusion
